% !TEX root = ../article.tex

% LaTeX template for Artifact Evaluation V20151015
%
% Prepared by Grigori Fursin (cTuning foundation, France and dividiti, UK) 
% and Bruce Childers (University of Pittsburgh, USA)
%
% (C)opyright 2014-2015


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% When adding this appendix to your paper, 
% please remove above part
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix
\section{Artifact Description}

%\newcommand{\nomcvm}{}

%Submission and reviewing guidelines and methodology: \\
%{\em http://cTuning.org/ae/submission-20151015.html}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Abstract}

\osrkit\ is a library that enables On-Stack Replacement (OSR) at arbitrary places in LLVM IR code. The artifact is designed to explore how \osrkit\ can instrument IR code to support OSR transitions in the LLVM MCJIT runtime environment. A running example is presented based on the \texttt{isord} case study discussed in Section 3. We also support repeating all the experiments presented in Section 5. The artifact includes an interactive VM called \tinyvm\ for loading, inspecting, instrumenting, and executing IR code. The package is a preconfigured Oracle VirtualBox VM.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Description}

The main component of the artifact is an interactive VM called \tinyvm\ built on top of the LLVM MCJIT runtime environment and the \osrkit\ library. The VM provides an interactive environment for IR manipulation, JIT-compilation, and execution of functions either generated at run-time or loaded from disk: for instance, it allows the user to insert OSR points in loaded functions, run optimization passes on them, display their CFGs, and repeatedly invoke a function for a specified amount of times. \tinyvm\ supports dynamic library loading and linking, and includes a helper component for MCJIT that simplifies tasks such as handling multiple IR modules, symbol resolution in presence of multiple versions of a function, and tracking machine-level generated code and data objects.

\tinyvm\ is located in {\small\tt /home/osrkit/Desktop/tinyvm/} and runs a case-insensitive command-line interpreter:
\begin{small}
\begin{verbatim}
osrkit@osrkit-AE:~/Desktop/tinyvm$ tinyvm
Welcome! Enter 'HELP' to show the list of commands.
TinyVM> 
\end{verbatim}
\end{small}

\noindent Use {\tt HELP} to print basic documentation on how to use the shell. Usage scenarios are discussed in \ref{ss:art-eval-res}.

\subsubsection{Check-list (artifact meta information)}

%{\em Fill in whatever is applicable with some informal keywords and remove the rest}

{\small
\begin{itemize}[parsep=0pt]
  %\item {\bf Algorithm: }
  \item {\bf Program: } {\tt shootout} C benchmarks (included, Sep 2015). %and a number of MATLAB benchmarks (included)
  \item {\bf Compilation: } LLVM 3.6.2 (release build).
  %\item {\bf Transformations: }
  %\item {\bf Binary: }
  %\item {\bf Data set: }
  \item {\bf Run-time environment: } Linux (version 3.x).
  \item {\bf Hardware: } x86-64 CPU.
  \item {\bf Run-time state: } Cache-sensitive (performance measurements only).
  %\item {\bf Execution: }
  \item {\bf Output: } Measures are output to console.
  \item {\bf Experiment workflow: } Invoke scripts and perform a few manual steps.
  \item {\bf Publicly available?} Yes.
\end{itemize}
}

\subsubsection{How Delivered}

The artifact ships as an Oracle VirtualBox 5 Appliance.
The latest version of the code is available at \url{https://github.com/dcdelia/tinyvm}.

\subsubsection{Hardware Dependencies}

An x86-64 platform is required.

\subsubsection{Software Dependencies}

The artifact was tested in Oracle VirtualBox 5.0.10. 

%\subsubsection{Datasets}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Installation}

To install the artifact, just import the appliance in Oracle VirtualBox, which installs Linux LXLE. Open the {\tt README} file on the {\tt Desktop} folder for further info on the artifact and the Linux distribution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment Workflow}

\ifdefined \nomcvm
We propose two usage sessions. In the first session, we show how to generate and instrument an LLVM IR code based on the \texttt{isord} example presented in \mysection\ref{se:osr-llvm}. The second session focuses on how to run the scripts used to generate the performance tables of \mysection\ref{se:experiments} related to questions Q1, Q2, and Q3. Question Q4 is based on additional third-party software (the MATLAB McVM runtime\footnote{The source code of the version used in the paper, which we ported to LLVM 3.6+ and extended with the {\tt feval} optimization technique discussed in \ref{ss:eval-opt-mcvm}, is available at \url{https://github.com/dcdelia/mcvm}.}) and is not addressed in the artifact.
\else
We propose three usage sessions. In the first session, we show how to generate and instrument an LLVM IR code based on the \texttt{isord} example presented in \mysection\ref{se:osr-llvm}. The second session focuses on how to run the scripts used to generate the performance tables of \mysection\ref{se:experiments} related to questions Q1, Q2, and Q3. The third session addresses question Q4, using third-party software (the MATLAB McVM runtime~\cite{mcvm}) that we ported to LLVM 3.6+ and extended with the {\tt feval} optimization technique discussed in \ref{ss:eval-opt-mcvm}.
\fi

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Evaluation and Expected Result}
\label{ss:art-eval-res}

\input{artifact/session1}
\input{artifact/session2}
\ifdefined \nomcvm
\else
\input{artifact/session3}
\fi

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Notes}

We encourage the reader to experiment with \tinyvm, creating IR programs with {\tt clang -S -emit-llvm -O1}, instrumenting them with OSR points, and exploring the generated code. Please bear in mind that \tinyvm\ is a prototype implementation that does not support exercising all the features for VM builders provided by \osrkit. For instance, while \osrkit\ is fully flexible, \tinyvm\ only supports always-firing and never-firing OSR points. Also, unexpected results may arise: we will be glad to hear about your experience and grateful to receive any bug reports.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% When adding this appendix to your paper, 
% please remove below part
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


